---
title: "Reproducible Pipelines with `targets`"
format:
  html:
    code-fold: false
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  echo = TRUE,
  comment = "#>",
  fig.path = "images/drafts-targets-"
)
```

# Introduction to `targets`

In data analysis projects, ensuring reproducibility is crucial. The `targets` package in R provides a powerful framework for building reproducible, scalable, and efficient data pipelines. This tutorial will guide you through the basics of using `targets` to manage your analysis workflow.

## Purpose

The primary purpose of using `targets` is to **prevent rerunning parts of the pipeline if their dependencies haven't changed**. This optimization saves time and computational resources by only executing the necessary steps when data or code modifications occur.

## Getting Started

### Installation

First, install the `targets` package from CRAN if you haven't already:

```{r}
install.packages("targets")
library(targets)
```

### Initialization

```{r}
#| include: false
if (file.exists("_targets.R")) {
  unlink("_targets.R")
}
if (dir.exists("_targets")) {
  unlink("_targets", recursive = TRUE)
}
```

Load the `targets` package and initialize a new `targets` pipeline in your project directory:

```{r}
#| eval: false
library(targets)
use_targets()
```

```{r}
#| include: false
# Read the file
file_path <- "_targets.R"
file_content <- readLines(file_path)

# Replace the specific line
file_content <- gsub("1000", "1e6", file_content)
file_content <- gsub("100", "1e6", file_content)

# Write the updated content back to the file
writeLines(file_content, file_path)
```

This command creates a `_targets.R` script, with helpful setup instructions.

Here is a slightly reduced version of it:

```r
# Created by use_targets().
# Follow the comments below to fill in this target script.
# Then follow the manual to check and run the pipeline:
#   https://books.ropensci.org/targets/walkthrough.html#inspect-the-pipeline

# Load packages required to define the pipeline:
library(targets)

# Set target options:
tar_option_set(
  packages = c("tibble") # Packages that your targets need for their tasks.
)

# Run the R scripts in the R/ folder with your custom functions:
tar_source()

# tar_source("other_functions.R") # Source other scripts as needed.

# Replace the target list below with your own:
list(
  tar_target(
    name = data,
    command = tibble(x = rnorm(1e4), y = rnorm(1e4))
    # format = "qs" # Efficient storage for general data objects.
  ),
  tar_target(
    name = model,
    command = coefficients(lm(y ~ x, data = data))
  )
)
```

## Running the pipeline

We then run this pipeline using `tar_make()`.
We do that below, and time how long the pipeline takes to execute:

```{r}
time_start <- proc.time()[3]
tar_make()
time_end <- proc.time()[3]
(time_end - time_start) |> signif(2)
```

Now, we can rerun it, and it will take less time:

```{r}
time_start <- proc.time()[3]
tar_make()
time_end <- proc.time()[3]
(time_end - time_start) |> signif(2)
```

It doesn't drive it down to zero as `targets` has some overhead (e.g., checking dependencies), but it's obviously faster.
This difference becomes dramatic when slow-running steps are involved.

If we adjust the pipeline, then `tar_make()` will rerun what is required:

```{r}
# Read the file
file_path <- "_targets.R"
file_content <- readLines(file_path)

# Replace the specific line
file_content <- gsub("y ~ x", "y ~ x - 1", file_content)

# Write the updated content back to the file
writeLines(file_content, file_path)

# Run the pipeline
time_start <- proc.time()[3]
tar_make()
time_end <- proc.time()[3]
(time_end - time_start) |> signif(2)
```

The time is still faster than before, as we didn't simulate the data, but it's slower than when the model fit was cached.

```{r}
#| include: false
# Read the file
file_path <- "_targets.R"
file_content <- readLines(file_path)

# Replace the specific line
file_content <- gsub("y ~ x - 1", "y ~ x", file_content)

# Write the updated content back to the file
writeLines(file_content, file_path)
```

## Accessing Outputs

After running the pipeline, you can access the outputs of the targets using:

- **`tar_load(<name>)`**: Loads the target object into the R environment.
  
  ```{r}
  tar_load(model)
  summary(model)
  ```
  
- **`tar_read(<name>)`**: Reads the target object without loading it into the environment.
  
  ```{r}
  model_summary <- tar_read(model)
  summary(model_summary)
  ```

## Debugging the Pipeline

Debugging is an essential part of developing a `targets` pipeline.
However, you cannot directly debug a `targets` pipeline by default (it runs the code "separately" from the R console).

Here are two methods to debug your pipeline.

### Option 1: Using `browser()`

Insert `browser()` into the function where you want to inspect the pipeline's state.

```{r}
#| eval: false
#| echo: true
tar_target(
  model,
  {
    browser()
    lm_model <- lm(value_log ~ predictor, data = data_processed)
    lm_model
  }
)
```

Run the pipeline with:

```{r}
targets::tar_make(callr_function = NULL)
```

This allows you to interactively debug the target execution.

### Option 2: Loading Everything

Load all targets into the environment to inspect and interact with them directly.

```{r}
targets::tar_load_everything()
```

This approach is useful for exploratory debugging and inspecting intermediate results.

For example, after running `targets::tar_load_everything()`, you can access the data object directly. You can then play around with the code to fit the model, having the data at hand. 

## More niche topics

### Multiple pipelines

Within a given project, one may define multiple pipelines.

### Forced Execution

Sometimes, you may want a target to **always rerun**, regardless of whether its dependencies have changed. Use `tar_cue(cue = "always")` to enforce this behavior.

```{r, eval=FALSE}
tar_target(
  figure,
  {
    plot <- ggplot(data_processed, aes(x = predictor, y = value_log)) +
      geom_point() +
      geom_smooth(method = "lm") +
      theme_minimal()
    ggsave("figures/model_plot.png", plot)
    plot
  },
  cue = tar_cue("always")
)
```

This ensures that the `figure` target is always executed when `tar_make()` is run.

Sometimes you just need to do this, e.g. `targets` can't pick up that the data has changed or the output has been wiped in between runs.

## Further Resources

- **Home Page**: [Dynamic Function-Oriented Make-Like Declarative Pipelines • targets](https://docs.ropensci.org/targets/)
- **Detailed User Manual**: [The {targets} R package user manual](https://books.ropensci.org/targets/)
